---
title: "Class 1 - Data Analysis in R"
author: "Tiago Afonso"
date: today
format: html
toc: true
---

# Class 1 - Data Analysis in `R`

### **Data Collection**
**WDI - World Development Indicators**


```{r}
#| label: load-libraries
#| output: false

library(tidyverse)
library(WDI)
```

Sample:

| Country | Code |
|---------|------|
| 🇩🇪 Germany | DEU |
| 🇫🇷 France | FRA |
| 🇧🇪 Belgium | BEL |

Series:

    `GDP (current US$)`
    `GDP deflator (base year varies by country)`
    `Tax revenue (% of GDP)`
    `Population, total`
    `Mobile cellular subscriptions (per 100 people)`

Time: 1990 - 2024

in `R`:

```{r}
#| label: load-data

wdi_data <- WDI(
  country = c("DEU", "FRA", "BEL"),
  indicator = c(
    "NY.GDP.MKTP.CD",  # GDP (current US$)
    "NY.GDP.DEFL.KD.ZG",  # GDP deflator (annual %)
    "GC.TAX.TOTL.GD.ZS",  # Tax revenue (% of GDP)
    "SP.POP.TOTL",  # Population, total
    "IT.CEL.SETS.P2"  # Mobile cellular subscriptions (per 100 people)
  ),
  start = 1990,
  end = 2024,
  extra = TRUE
) |>
  rename(
    gdp_current_usd = NY.GDP.MKTP.CD,
    gdp_deflator_annual_pct = NY.GDP.DEFL.KD.ZG,
    tax_revenue_pct_gdp = GC.TAX.TOTL.GD.ZS,
    population_total = SP.POP.TOTL,
    mobile_subscriptions_per_100 = IT.CEL.SETS.P2
  )

head(wdi_data)
summary(wdi_data)
```

The function `WDI()` downloads the data directly from the World Bank API. The function `rename()` is used to rename the columns.

Cleaning data:

- Filter for relevant years only (1995-2022)
- Select relevant columns (`country`, `year`, `gdp_current_usd`, `gdp_deflator_annual_pct`, `population_total`, `tax_revenue_pct_gdp`, `mobile_subscriptions_per_100`)


```{r}
#| label: clean-data


wdi_data_clean <- wdi_data |>
  filter(year >= 1995 & year <= 2022) |>
  select(country, year, gdp_current_usd, gdp_deflator_annual_pct, population_total, tax_revenue_pct_gdp, mobile_subscriptions_per_100)

head(wdi_data_clean)
summary(wdi_data_clean)
```

### **Exercise 1** 

Calculate the deflator for the base year **2015** for each country

Change the deflator base year to from 2020 to **2015**:

$$
Deflator_{2015} = \frac{Deflator_{t}}{Deflator_{2015}} \times 100
$$

where t=1995,...,2022

```{r}
#| label: exercise-1

wdi_data_clean <- wdi_data_clean |>
  group_by(country) |>
  mutate(
    gdp_deflator_base_2015 = (gdp_deflator_annual_pct / gdp_deflator_annual_pct[year == 2015]) * 100
  ) |>
  ungroup()
head(wdi_data)
```

The `mutate()` function is used to create a new columns in a data frame. The `group_by()` function is used to group the data by country, so that the calculations are done for each country separately. The `ungroup()` function is used to remove the grouping structure.

### **Exercise 2** 

Calculate the GDP at **2015** prices for each country

Change the GDP current to 2015 prices:

$$
GDP_{2015} = \frac{GDP_{current}}{Deflator_{2015}} \times 100
$$

where t=1995,...,2022

```{r}
#| label: exercise-2

wdi_data_clean <- wdi_data_clean |>
  mutate(
    gdp_2015_usd = (gdp_current_usd / gdp_deflator_base_2015) * 100
  )
```

### **Exercise 3** 
Chart the evolution of GDP per capita variation and subscriptions over time *(for one country only)*

1. Calculate GDP per capita at 2015 prices

$$
GDP_{per capita 2015} = \frac{GDP_{2015}}{Population_{total}}
$$


```{r}
#| label: select France

wdi_data_clean <- wdi_data_clean |>
  mutate(
    gdppc = gdp_2015_usd / population_total
  )

wdi_data_france <- wdi_data_clean |>
  filter(country == "France")
```

2. Calculate the annual variation of GDP per capita at 2015 prices

$$
\Delta GDPpc = GDPpc_{t} - GDPpc_{t-1}
$$

where t=1996,...,2022

```{r}
#| label: exercise-3

wdi_data_france <- wdi_data_france |>
  arrange(year) |>
  mutate(
    gdppc_var = gdppc - lag(gdppc)
  )
```

3. Chart the evolution of GDP per capita variation and subscriptions over time for France

```{r}
#| label: plot-exercise-3

# Plot 1: GDP per capita variation over time
ggplot(wdi_data_france, aes(x = year, y = gdppc_var)) +
    geom_line(color = "blue", size = 1) +
    geom_point(color = "darkblue", size = 4) +
    labs(
        title = "GDP per capita variation over time - France",
        x = "Year",
        y = "GDP per capita variation (USD)"
    ) +
    theme_minimal()

# Plot 2: Mobile subscriptions per 100 people over time
ggplot(wdi_data_france, aes(x = year, y = mobile_subscriptions_per_100)) +
    geom_line(color = "red", size = 1) +
    geom_point(color = "darkred", size = 4) +
    labs(
        title = "Mobile subscriptions per 100 people over time - France",
        x = "Year",
        y = "Mobile subscriptions per 100 people"
    ) +
    theme_minimal()
```

### **Exercise 4** 

Is there any relationship between GDP per capita and subscriptions? *(for each country)*

1. Using a scatter plot, for each country, plot GDP per capita at 2015 prices against mobile subscriptions per 100 people.

```{r}
#| label: plot-exercise-4

ggplot(wdi_data_clean, aes(x = mobile_subscriptions_per_100, y = gdppc)) +
    geom_point(aes(color = country), size = 2) +
    labs(
        title = "GDP per capita vs Mobile subscriptions per 100 people",
        x = "Mobile subscriptions per 100 people",
        y = "GDP per capita (2015 prices)"
    ) +
    theme_minimal()
```

Adding a linear regression to each country:

```{r}
#| label: plot-exercise-4-regression

ggplot(wdi_data_clean, aes(x = mobile_subscriptions_per_100, y = gdppc)) +
    geom_point(aes(color = country), size = 2) +
    geom_smooth(aes(color = country), method = "lm", se = FALSE) +
    labs(
        title = "GDP per capita vs Mobile subscriptions per 100 people with regression lines",
        x = "Mobile subscriptions per 100 people",
        y = "GDP per capita (2015 prices)"
    ) +
    theme_minimal()
```

If flat, no relationship. If upward sloping, positive relationship. If downward sloping, negative relationship.

Answering the same question using correlation (pearson correlation):

```{r}
#| label: correlation-exercise-4

wdi_data_clean %>%
  group_by(country) %>%
  summarize(correlation = cor(gdppc, mobile_subscriptions_per_100, use = "complete.obs"))
```

The `cor()` function is used to calculate the correlation between two variables. The `use = "complete.obs"` argument is used to handle missing values by only using complete observations.


### **Exercise 5**

Calculate the two statistical moments for all variables *(France only)*

```{r}
#| label: moments-exercise-5

#unselect the  country and year columns

wdi_data_france_ <- wdi_data_france %>%
  select(-country, -year)

library(psych)

describe(wdi_data_france_)
```

Check for the existence of **outliers** for each variable *(France only)*

#### **Formulae for Outlier Detection**

$$
{IQR = Q3 - Q1}
$$

$$
{Lower\;Limit = Q1 - 1.5 \times IQR}
$$

$$
{Upper\;Limit = Q3 + 1.5 \times IQR}
$$

```{r}
#| label: outliers-exercise-5

wdi_data_france_outliers <- wdi_data_france_ %>%
  gather(key = "variable", value = "value")
head(wdi_data_france_outliers)
wdi_data_france_outliers_summary <- wdi_data_france_outliers %>%
  group_by(variable) %>%
  summarize(
    Q1 = quantile(value, 0.25, na.rm = TRUE),
    Q3 = quantile(value, 0.75, na.rm = TRUE),
    IQR = IQR(value, na.rm = TRUE),
    Lower_Limit = Q1 - 1.5 * IQR,
    Upper_Limit = Q3 + 1.5 * IQR
  )
wdi_data_france_outliers_summary
```

If values are below the Lower Limit or above the Upper Limit, they are considered outliers.

```{r}

# Identify outliers
wdi_data_france_outliers <- wdi_data_france_outliers %>%
  left_join(wdi_data_france_outliers_summary, by = "variable") %>%
  mutate(
    is_outlier = ifelse(value < Lower_Limit | value > Upper_Limit, TRUE, FALSE)
  )

#count number of outliers for each variable
wdi_data_france_outliers %>%
  group_by(variable) %>%
  summarize(num_outliers = sum(is_outlier, na.rm = TRUE))
 
```

Boxplot for each variable *(France only)*


```{r}
#| label: boxplots-exercise-5

library(gridExtra)

# Create individual boxplots for each variable with colors
p1 <- wdi_data_france %>%
    ggplot(aes(x = "", y = gdp_current_usd)) +
    geom_boxplot(fill = "lightblue", color = "darkblue") +
    labs(title = "GDP Current USD", x = "", y = "Value") +
    theme_minimal()

p2 <- wdi_data_france %>%
    ggplot(aes(x = "", y = gdp_deflator_annual_pct)) +
    geom_boxplot(fill = "lightgreen", color = "darkgreen") +
    labs(title = "GDP Deflator Annual %", x = "", y = "Value") +
    theme_minimal()

p3 <- wdi_data_france %>%
    ggplot(aes(x = "", y = population_total)) +
    geom_boxplot(fill = "lightcoral", color = "darkred") +
    labs(title = "Population Total", x = "", y = "Value") +
    theme_minimal()

p4 <- wdi_data_france %>%
    ggplot(aes(x = "", y = tax_revenue_pct_gdp)) +
    geom_boxplot(fill = "lightyellow", color = "orange") +
    labs(title = "Tax Revenue % GDP", x = "", y = "Value") +
    theme_minimal()

p5 <- wdi_data_france %>%
    ggplot(aes(x = "", y = mobile_subscriptions_per_100)) +
    geom_boxplot(fill = "lightpink", color = "purple") +
    labs(title = "Mobile Subscriptions per 100", x = "", y = "Value") +
    theme_minimal()

p6 <- wdi_data_france %>%
    ggplot(aes(x = "", y = gdppc)) +
    geom_boxplot(fill = "lightgray", color = "black") +
    labs(title = "GDP per Capita", x = "", y = "Value") +
    theme_minimal()

# Arrange plots in a grid
grid.arrange(p1, p2, p3, p4, p5, p6, ncol = 3, nrow = 2)
```


With another package:

```{r}

library(plotly)

p1 <- plot_ly(y = wdi_data_france$gdp_current_usd, type = "box", name = "GDP Current USD", boxpoints = "all", jitter = 0.3, pointpos = -1.8, marker = list(color = 'darkblue'), line = list(color = 'blue'))
p2 <- plot_ly(y = wdi_data_france$gdp_deflator_annual_pct, type = "box", name = "GDP Deflator Annual %", boxpoints = "all", jitter = 0.3, pointpos = -1.8, marker = list(color = 'darkgreen'), line = list(color = 'green'))
p3 <- plot_ly(y = wdi_data_france$population_total, type = "box", name = "Population Total", boxpoints = "all", jitter = 0.3, pointpos = -1.8, marker = list(color = 'darkred'), line = list(color = 'red'))
p4 <- plot_ly(y = wdi_data_france$tax_revenue_pct_gdp, type = "box", name = "Tax Revenue % GDP", boxpoints = "all", jitter = 0.3, pointpos = -1.8, marker = list(color = 'orange'), line = list(color = 'yellow'))
p5 <- plot_ly(y = wdi_data_france$mobile_subscriptions_per_100, type = "box", name = "Mobile Subscriptions per 100", boxpoints = "all", jitter = 0.3, pointpos = -1.8, marker = list(color = 'purple'), line = list(color = 'pink'))
p6 <- plot_ly(y = wdi_data_france$gdppc, type = "box", name = "GDP per Capita", boxpoints = "all", jitter = 0.3, pointpos = -1.8, marker = list(color = 'black'), line = list(color = 'gray'))

subplot(p1, p2, p3, p4, p5, p6, nrows = 2, margin = 0.05) %>%
  layout(title = "Boxplots for France Variables")
```